---
title: "FoB problems"
author: "Yigit Ozan Berk"
date: "8/22/2019"
output: html_document
---

SMOKE.dat

```{r}
library(miceadds)
library(dplyr)
load.Rdata(filename = "./data/SMOKE.DAT.rdata", "smoke.dat")
head(smoke.dat)
```

```{r}
df1 = mutate(smoke.dat, month = ntile(Day_abs, 12))
head(df1)
```

```{r}
library(ggplot2)
qplot(df1$month, fill = I("magenta"))
```

```{r}
probs = df1 %>% group_by(month) %>% summarise(n(), prob = n()/nrow(df1))
probs
```

```{r}
probs[c(1, 3, 6), ]
```

```{r}
#initial model
fit1 = glm(Day_abs ~ Cig_day, family = poisson, data = df1)
fit2 = glm(Day_abs ~ Age + Cig_day, family = poisson, data = df1)
fit3 = glm(Day_abs ~ Age + I(as.factor(Gender)) + Cig_day, family = poisson, data = df1)
fit4 = glm(Day_abs ~ Age + I(as.factor(Gender)) + Cig_day  + co, family = poisson, data = df1)
summary(fit1)$coef
summary(fit2)$coef
summary(fit3)$coef
summary(fit4)$coef
anova(fit1, fit2, fit3, fit4)
```

4th model fits best apparently. the beginnig co levels in blood seems to be very significant for quitting.

```{r}
summary(fit4)

```

```{r}
exp(confint(fit4))
```

```{r}
1- exp(summary(fit4)$coef[5,1])
```

Increase of 1 point in starting CO levels in blood results in .119 % decrease in days abstinent.
```{r}
exp(summary(fit4)$coef[1,1]) -1
exp(summary(fit4)$coef[2,1]) -1
exp(summary(fit4)$coef[3,1]) -1
exp(summary(fit4)$coef[4,1]) -1
```

The average days abstinent for average age, average cigs_day and average co levels for males is 71.

females tend to increase abstinent days 3.81%.


```{r}
data(sleep)
sleep
```

everything is normalized according to 10 hours

```{r}
g1 = sleep$extra[1:10]
g2 = sleep$extra[11 : 20]
difference = g2 - g1
mn = mean(difference) # 1.67
s = sd(difference) #1.13
n = 10
mn + c(-1, 1) * qt(.975, n-1) * s / sqrt(n) #95 % conf. int.
t.test(difference)$conf.int
```



