---
title: "Regression Modelling Week 2"
author: "Yigit Ozan Berk"
date: "7/10/2019"
output: html_document
---

# Statistical Linear Regression Models

Up to this point, we’ve only considered estimation. Estimation is useful, but we also need to know how to extend our estimates to a population. This is the process of statistical inference. Our approach to statistical inference will be through a statistical model. At the bare minimum, we need a few distributional assumptions on the errors. However, we’ll focus on full model assumptions under Gaussianity.


notes on the paper sheet

## linear regression for prediction

example from 'diamond' data set from UsingR

```{r}
library(UsingR)
data(diamond)
library(ggplot2)
g <- ggplot(diamond, aes(x = carat, y = price))
g <- g + xlab("Mass (carats)")
g <- g + ylab("Price (SIN $)")
g <- g + geom_point(size = 6, colour = "black", alpha = 0.2)
g <- g + geom_point(size = 5, colour = "blue", alpha = 0.2)
#black background and blue top colour
g <- g + geom_smooth(method = "lm", colour = "black")
#assumes y as outcome, x as predictor
g

```

```{r}
fit <- lm(price ~ carat, data = diamond)
coef(fit)
```

we estimate an expected `r round(coef(fit)[2], 2)` (SIN) $ for every carat increase in mass of diamond.

```{r}
summary(fit)
```

```{r}
#mean centering
fit2 <- lm(price ~ I(carat - mean(carat)), data = diamond)
#if you want mathmatical functions within lm function, you have to surround them in the I() function
coef(fit2)
```

500 SIN $ is the average price for the average sized diamond which is about X = 0.2.

```{r}
#changing units of slope to 1/10 of a carat
fit3 <- lm(price ~ I(carat * 10), data = diamond)
coef(fit3)
```

```{r}
#predicting the price
newx <- c(0.16, 0.27, 0.34)
coef(fit)[1] + coef(fit)[2]*newx

```
```{r}
predict(fit, newdata = data.frame(carat = newx))
#takes output from several different prediction models
# if you don't use the newdata variable, it predicts yhat values of the original x values in the model 'fit'
```



